Demes: a specification for interchanging complex demography

### 2022-04-20 paper.tex
address comments and edits throughout

\aprcomment{Note: Need to standardize use of ``demes'' vs ``Demes'' vs \Demes.}
\section*{Introduction}

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales.
An integral component to such population genetics studies is simulation.
Software to either simulate whole genome
sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
or informative summary statistics of
diversity~\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
have enabled the increasing complexity of genomic studies, with several software
packages capable handling large sample sizes, many interacting populations, and
deviations from panmictic random-mating assumptions.
This ability to infer and simulate such complex demographic scenarios, however,
has highlighted a major shortcoming in community standards.
The fragmented landscape of different approaches to describing demographic
models makes it difficult to compare inferences made by different methods
and to reliably simulate from previously inferred models.

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information. 
\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
However, there is no standard format shared across simulators,
and the details of the demographic models are not separated
from other simulation parameters.
The APIs used by individual simulators to define demographic models are also
not appropriate as an interchange standard because of the dependency on both
the programming language involved,
and the specific details of the simulator in question.

@@

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
    \begin{center}
        \begin{tabular}{lp{12cm}}
            \softwaretable
        \end{tabular}
    \end{center}
    \caption{
        \label{tab:software}
        \textbf{Software support for Demes.}
        We have included software infrastructure developed
        for working with Demes models (parsing, validation,
        visualisation, etc) as well as all known downstream 
        software that implements the specification,
        at the time of writing.
    }
\end{table}

@@

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The deomgraphic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\aprcomment{citations here, and revisit} or
\aprcomment{stochastic carrying-capacity models, or something like that},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
\jkcomment{I guess these are also the basic assumptions also shared by
inference methods, which is important. Mabye another sentence here making this
explicit, and listing a few different types of inference methods and giving
citations? This is critical - we want to establish that Demes is the
``middleware'' between the inference methods and the simulators, and captures the
basic assumptions made by them all.}
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

\aprcomment{redundancies with Appendix - that could be expanded and reworked a bit}

@@

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{fig/showcase}
    \caption{
        \textbf{Illustration and simulation using Demes.}
        (A)
        Using an inferred demographic model from \citet{tennessen2012evolution}
        specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
        use \demesdraw\ to visualize the demographic model.
        We then use \msprime\ to simulate genomic data for 20 genome copies
        sampled from the two contemporary populations, and we use \moments\
        to compute the expected joint site-frequency spectrum for the same
        sample sizes (Figure~\ref{fig:tennessen-simulation}).
        (B, C) We compare the single-population SFS in each population, showing
        agreement between the simulation methods.
        (D) Code snippets of the interactions between \demes\ and the simulation
        software. An extended script to compute the SFS shown in (B) and (C) is
        given in Figure~\ref{fig:tennessen-simulation}.
        \aprcomment{There may be a better model to use here -- the massive
        exponential growth really makes the rest of the history hard to see.}
    }
    \label{fig:showcase}
\end{figure}

@@

We used this model to simulate 20 haploid genome copies from each EUR and AFR
at time zero (i.e., present day) to obtain the joint site-frequency spectrum
(SFS), a summary of observed allele frequencies widely used in evolutionary
inference
\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
The Demes model (Figure~\ref{fig:showcase}A~and~\ref{fig:tennessen}) was
provided as the input demography to \msprime\ \citep{baumdicker2021-iu} to
simulate a large recombining region with the mutation rate assumed in
\citet{tennessen2012evolution}, and we computed the observed SFS using \tskit\
\citep{ralph2020efficiently}. Using the same Demes model as input to \moments\
\citep{jouganous2017inferring}, we computed the expectation of the joint SFS
and compared to the \msprime\ simulated data (Figure~\ref{fig:showcase}B,C).
To reiterate, for both \msprime\ and \moments\ simulations, the same
demographic model was used as input, avoiding the need to specify the complex
demographic model multiple times using the unique API of each software.

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
ewing2010msms,zhou2018popdemog}. However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a front end for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

@@

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161} and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardisation in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by simulation and inference methods and be used for reporting results in
publications, either as supplemental material or uploaded to a data
repository in Demes format.

@@

The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task and can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

>>>>

\section*{Introduction}

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales.
An integral component to such population genetics studies is simulation.
Software to either simulate whole genome
sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,
baumdicker2021-iu,kelleher2016efficient,haller2019slim}
or informative summary statistics of
diversity~\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
have enabled the increasing complexity of genomic studies, with several software
packages capable handling large sample sizes, many interacting populations, and
deviations from panmictic random-mating assumptions.
This ability to infer and simulate such complex demographic scenarios, however,
has highlighted a major shortcoming in community standards.
The fragmented landscape of different approaches to describing demographic
models makes it difficult to compare inferences made by different methods
and to reliably simulate from previously inferred models.

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information. 
% \mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
% APR: I think this is ok here.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.,][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.,][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
However, there is no standard format shared across simulators,
and the details of the demographic models are not separated
from other simulation parameters.
The APIs used by individual simulators to define demographic models are also
not appropriate as an interchange standard because of the dependency on both
the programming language involved
and the specific details of the simulator in question.

@@

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
    \begin{center}
        \begin{tabular}{lp{12cm}}
            \softwaretable
        \end{tabular}
    \end{center}
    \caption{
        \label{tab:software}
        \textbf{Software support for Demes.}
        We have included software infrastructure developed
        for working with Demes models (such as parsing, validation,
        and visualisation) as well as downstream 
        software that implement the specification,
        at the time of writing.
    }
\end{table}

@@

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The demographic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\citep{wright1943isolation,barton2002neutral,ringbauer2017inferring,battey2020space}
or stochastic population size dynamics
\citep{nunney1993assessing,orr2014population},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
The basic assumptions of discrete populations connected by instantaneous or continuous
migrations are also shared by many inference methods
\citep[e.g.,][]{gutenkunst2009inferring,gravel2012population,kamm2017efficient,
jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
Demes therefore serves as ``middleware'' between inference methods and simulation
software, capturing common assumptions shared across implemented software.
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

@@

\aprcomment{I think this belongs here, just one or two sentences:}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{fig/showcase}
    \caption{
        \textbf{Illustration and simulation using Demes.}
        (A)
        Using an inferred demographic model from \citet{tennessen2012evolution}
        specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
        used \demesdraw\ to visualize the demographic model (note the recent
            exponential growth resulting in present-day population sizes that
        greatly exceed those in the past).
        We then used \msprime\ to simulate genomic data for 20 genome copies
        sampled from the two contemporary populations, and we used \moments\
        to compute the expected joint site-frequency spectrum for the same
        sample sizes (Figure~\ref{fig:tennessen-simulation}).
        (B, C) We compared the single-population SFS in each population, showing
        agreement between the simulation methods.
        (D) Code snippets of the interactions between \demes\ and the simulation
        software. An extended script to compute the SFS shown in (B) and (C) is
        given in Figure~\ref{fig:tennessen-simulation}.
    }
    \label{fig:showcase}
\end{figure}

@@

We used this model to simulate 20 haploid genome copies from each EUR and AFR
at time zero (i.e., present day) to obtain the joint site-frequency spectrum
(SFS), a summary of observed allele frequencies widely used in evolutionary
inference
\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,
jouganous2017inferring,kamm2017efficient,kim2017inference}.
The Demes model (Figures~\ref{fig:showcase}A~and~\ref{fig:tennessen}) was
provided as the input demography to \msprime\ \citep{baumdicker2021-iu} to
simulate a large recombining region under the mutation rate assumed in
\citet{tennessen2012evolution}, and we computed the observed SFS using \tskit\
\citep{ralph2020efficiently}. Using the same Demes model as input to \moments\
\citep{jouganous2017inferring}, we computed the expectation of the joint SFS
and compared to the \msprime\ simulated data (Figure~\ref{fig:showcase}B,C).
To reiterate, for both \msprime\ and \moments\ simulations, the same
demographic model was used as input, avoiding the need to specify the complex
demographic model multiple times using the unique API of each software.

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[e.g.,][]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
ewing2010msms,zhou2018popdemog}. However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a front end for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

@@

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161}, and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardisation in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by simulation and inference methods and be used for reporting results in
publications, either as supplemental material or uploaded to a data
repository in Demes format.

@@

The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task and can be shared by programs that
support Demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.
